---
title: "Converting NTSB database downloads"
author: "Todd Curtis"
date: "December 12, 2015"
output: html_document
---

###Summary###
The National Transportation Safety Board (NTSB) maintains a database of aviation accidents and incidents that can be accessed by the general public on the NTSB web site. That  database contains information about accidents and selected incidents within the United States, its territories and possessions, and in international waters. The database also includes events involving US-registered aircraft that occur outside US territory.

While the database is publicly accessible on the NTSB web site, users are limited in how the information can be used. The records can be displayed on the web site, or the results of a search may be downloaded. However, because the downloads are in the form of a text or XML file, rather than a file type that can be used by common spreadsheet or database programs, users must first transform the data into a form that can be analyzed, but no resources are provided by the NTSB for that purpose.

While many spreadsheet programs, including Excel and LibreOffice, may be able to covert a text or an XML file into a common format such as a CSV file (particularly the NTSB text file format, which uses the vertical bar (|) as delimiter), the data may still have to be transformed in other ways in order to be analyzed by R or other data analysis programs.

Using the data analysis program R, AirSafe.com has created two programs that can take either the text file or XML file version of the output and turn it into a CSV file that can be used by widely available spreadsheet and data analysis programs. AirSafe.com has made the following resources available to the public, the R programs that perform either a text to CSV or XML to CSV file transformation, and a CSV file representing the download of the entire database of over 77,000 records from 1982 to 2015.

###Background###
The National Transportation Safety Board is an independent US government agency that has as one of its missions the investigation of accidents involving air transportation. As part of its mission, it provides a wide range of aviation safety information to the general public, including access to a database of aircraft accidents and incidents. The database, which contain records from 1982 to the present, can be accessed at [http://www.ntsb.gov/_layouts/ntsb.aviation/index.aspx](http://www.ntsb.gov/_layouts/ntsb.aviation/index.aspx). Each record has a variety of data about each incident and accident, including the date and location of the event, and the number and type of injuries. 

Users can choose to search for specific event characteristics, and can have the results either displayed in their browser or downloaded into either an XML file called AviationData.xml or a text file called AviationData.txt. Users also have the option of downloading the entire database of over 77,000 records.

While the NTSB provides various data summaries elsewhere on the site and in reports released to the public, the organization does not provide either any resources for users to create their own summaries using resources on the site, nor does it provide any instructions on how to transform the data into a form that could be used by a spreadsheet program or other data analysis program. 

###Objective###
AirSafe.com, which since 1996 has provided the public with extensive information and resources related to airline safety and security, sought to create a process that could both transform the NTSB database output into a form that could be analzyed by AirSafe.com, and to provide other researchers with the resources that could allow them to conduct their own analyses. Those resources include the following:

* Two programs, each written in the statistical computing language R, that would allow a user to transform an XML or text file output from the NTSB online database into into a data format that could be analyzed by R, or into a CSV file that could be used by other statistical computing programs or even spreadsheet programs such as Excel or Google Sheets.

* A data dictionary that describes each of the variables created by the AirSafe.com program.

* A description of the steps needed to transform the NTSB data.

* Examples of the kinds of statistical analyses, including graphic presentations, that could be accomplished using R.

The last two resources are covered in detail in this report, and there are links to the other resources at the end of this document.

###Who would find this information useful?###
There are several kinds of groups or individuals who may find some or all of the above resources useful. Some of those individuals or groups include the following:

* Current users of the NTSB online database who have a need or a desire to anaylze the database using tools that are not provided on the NTSB web site.

* Aviation professionals or organizations of aviation professionals who want a better understanding of the historical risks associated with their professions.

* Organizations that currently rely on the summary aviation safety statistics created by the NTSB, and that may need to create customized summary statistics.

* Journalists and news media organizations.

* Data scientists or data science students who are seeking out authentic and publicly available data for teaching, training, or research purposes.

###How to get the most out of this report###
There are three ways that the information associated with this report will likely be used:

1. The CSV data and the data dictionary associated with this report may be taken and used to perform analyses with a program other than R. To do this, the reader can simply go to the end of this report to find links to the CSV data (representing the entire contents of the NTSB database, encompassing the period from January 1982 to early December 2015).

2. The R programs developed for this report may be used to transform the NTSB database outputs for further analysis in R. Links to both programs are in the Resources section at the end of this report. If the user will only need to analyze NTSB data from earlier than the first week of December 2015, the CSV file in the Resources section already has all of that information.

3. The R programs associated with this report may be used as a template or stepping stone for a revised R program, or to develop similar programs for another data analysis programming environment. Links to a pair of programs, one to transform the XML NTSB data output and one to transform the text file output, are in the resources section. Both programs also contain sample summary analyses and data graphics to help illustrate what can be done with the processed data.

All three kinds of users would find the data dictionary useful, as it provides detailed descriptions of the variables used in the NTSB database, as well as the variables that were added by AirSafe.com to make it easier to create useful summary statistics from the NTSB data.

While the first and second types of users can proceed to Summary Statistics and Resources section at the end of this document, the third type of user who intends to alter the R programs in the Resources section or who wish to create a similar program in another programming language should review the rest of this document.

###Transformation programs###
Both the R program to transform a text NTSB output file and the one to transform an XML NTSB output file went through 10 general transformation steps (Step 0 to Step 9) to create a downloadable CSV file. Eight of these steps (Step 2 to Step 9) were the same in both programs, and the differences in the first two steps (Step 0 and Step 1) will be noted below. 

####Transformation procedures####
The transformation process assumes three things:

1. That the user has installed the R programming language in their computer, 
2. That the user has the appropriate R program in their working directory in that computer, and 
3. That the user has downloaded into the same working directory either the text file or XML file associated with a search on the NTSB online database.

#####Step 0: Load the NTSB output file #####
The text conversion program uploads the file "AviationData.txt" from the same working directory of the R program. Note that when downloading the output from the NTSB web site, it may save the file under a different name and in a location other than your working directory.

The XML conversion program assumes the file has the name "AviationData.xml"

The text file conversion program would removes the field separator character and transforms the variable names into ones appropriate for a data frame (replacing any spaces in the variable name with a period). For example, the variable name "Investigation Type" becomes "Investigation.Type". 

##### Step 1: Eliminate extra columns (text) or specify variable names (XML)#####
The text file conversion process adds a column with is devoid of data in the data frame, and in the text file program this column is removed. 

The XML conversion program does not add any extra columns. Instead, Step 1 in this program ensures that the variable names used are the same as the type used in the text conversion program.

#####Step 2: Ensure character variables are of type character #####
Variables identified from the NTSB data dictionary as being of type character are specified as such in the R program. Also in this step, all data from character variables are checked, and any leading or trailing space characters are removed. Space characters include tab, newline, vertical tab, form feed, carriage return, space.

#####Step 3: Ensure numerical variables are of type numeric#####
Variables identified from the NTSB data dictionary as being of type numeric are specified as such in the R program. This step also replaces any missing or blank numeric values with the symbol NA, which stands for "Not Available".

#####Step 4: Change date variables into a format suitable for R#####
The NTSB database has two date variables, one for the date of the event and a second for the date of the publication of the accident or incident report. Dates are in the date format dd/mm/YYYY, and are converted to the date format YYYY-mm-dd. 

This step also takes the date of the event and adds three more variables: the year of the event, the month the event occurred, and the day of the week of the event's occurrence.

#####Step 5: Eliminate any row (record) which has no date#####
Any event with a missing value for the date of the event are removed and not used in any subsequent analysis.

#####Step 6: Change blank, "N/A","Unknown", and similar responses to NA#####
All missing data is replaced with the R symbol NA. Missing data in an event's record could be be blank, which would occur for a numeric variable, or for character variables it could match one of the following character strings: "", ".", "N/A", "UNK","Unknown",or "Unavailable".

#####Step 7: Specify city name, state abbreviation, and full state name#####
The NTSB uses a location variable that typically includes the city and two-letter state abbreviation for any US location. The program adds three additional variables for the US city, two-letter state abbreviation, and full state name. It then scans the location variable for any character string that includes at least one comma, checks to see if the last comma is followed by a two-character sequence that matches a state code. If that is the case, the city name, state abbreviation, and full state name are added to those events.

A review of the full database revealed that for US locations, two-character abbreviations associated with standard US Postal Service codes were used for the 50 US states, the District of Columbia, and other US territories. Also, there were several dozen instances where non-standard two-character codes were used for locations in the Gulf of Mexico (GM), Atlantic Ocean (AO), and Pacific Ocean (PO). 

There were a total of 62 two-character identifiers used in the program, with 59 representing standard US Postal Service identifiers for states, territories, and the District of Columbia; as well as three non-standard codes for bodies of water.

#####Step 8: Arrange the new columns in logical groupings#####
The new variables (columns) associated with  the date and location of the event were appended to the end (right side) of the R data frame. The columns representing the new variables were regrouped so that they were adjacent to the appropriate NTSB variable.

#####Step 9: Save the processed data frame as a CSV file#####
The last step saves the output into your working directory in a CSV file named "ntsb_data.csv".

###Sample summary statistics###
The following are a sample of the kind of summary statistics and graphics that can be generated using the R programming language. They are based on a download of the complete database that was made on 12 December 2015.
```{r echo=FALSE}
filename <- "AviationData.txt" 
ntsb.data.raw = read.delim(filename, header=TRUE, sep="|",  quote = "",dec = ".",fill = TRUE, comment.char = "")
ntsb.data = ntsb.data.raw

# DATA CLEANING

# Step 1: Eliminate any columns (variables) added as a consequence of Step 0,
# columbns which have no data (all are NA)
elim.var = which(apply(ntsb.data, 2, function(x) all(is.na(x))))
elim.var = as.data.frame(elim.var) # Column name and column number of this data frame are the columns to be cut

# Eliminates all columns with only NA values
if (nrow(elim.var)>0) ntsb.data = ntsb.data[,-elim.var[,1]] 

# Step 2: Ensure character variables are of type character

# As part of the process, need to ensure that there are not extra spaces
# at the beginning or end of each character value

# ================================
# Use function 'stripper', which works like the str_trim() function in the strigr package  
# Using this function to use only base R package.

stripper <- function(x){
        # This function removes leading and trailing spaces from a vector.
        # Space characters include tab, newline, vertical tab, form feed, 
        # carriage return, space.
        # Equivalent to the str_trim() function in the strigr package   
        x = as.character(x)
        x = sub("[[:space:]]+$", "", x) # Remove leading space characters
        x = sub("^[[:space:]]+", "", x) # Remove trailing space characters
        return(x)
}
# ================================

# Find which columns match the following known character vectors
char.vecs = c("Event.Id","Investigation.Type","Accident.Number",       
              "Location", "Country", "Airport.Code", "Airport.Name", "Injury.Severity",
              "Aircraft.Damage", "Aircraft.Category", "Registration.Number", "Make",
              "Model", "Amateur.Built", "Engine.Type", "FAR.Description","Schedule",              
              "Purpose.of.Flight", "Air.Carrier", "Weather.Condition",
              "Broad.Phase.of.Flight", "Report.Status")

char.vecs.ndx = which(colnames(ntsb.data) %in% char.vecs)

# Ensure that the character variable is of type character, then remove extra spaces
for (i in 1:length(char.vecs.ndx)) {
        ntsb.data[,char.vecs.ndx[i]] = as.character(ntsb.data[,char.vecs.ndx[i]])
        ntsb.data[,char.vecs.ndx[i]] = stripper(ntsb.data[,char.vecs.ndx[i]])
}

# Step 3: Ensure numerical variables are of type numeric

# Find which columns match the following known numerical vectors
num.vecs = c("Latitude", "Longitude", "Number.of.Engines", "Total.Fatal.Injuries",  
             "Total.Serious.Injuries", "Total.Minor.Injuries", "Total.Uninjured")

# Creates a vector for the column numbers for numeric variables
num.vecs.ndx = which((colnames(ntsb.data)) %in% num.vecs)

# Note: This step appears to replace missing numeric values with NA
for (i in 1:length(num.vecs.ndx)) {
       ntsb.data[,num.vecs.ndx[i]] = as.numeric(as.character(ntsb.data[,num.vecs.ndx[i]]))
}


# Step 4: Change date variables into a format suitable for R

# Dates are in form mm/dd/YYYY, must convert to a date format of YYYY-mm-dd
# Two date variables, Event.Date and Pulication Date
ntsb.data$Event.Date = as.Date(ntsb.data$Event.Date, "%m/%d/%Y")
ntsb.data$Publication.Date = as.Date(ntsb.data$Publication.Date, "%m/%d/%Y")
# Note: This step appears to replace missing date values with NA

# Now will have separate columns for Year, Month, Day, and Weekday for Event.Date
ntsb.data$Year = as.numeric(format(ntsb.data$Event.Date, "%Y")) # Ensure it is a numeric variable
ntsb.data$Month = months(ntsb.data$Event.Date, abbreviate = TRUE)
ntsb.data$Day = format(ntsb.data$Event.Date, "%d")
ntsb.data$Weekday = weekdays(ntsb.data$Event.Date, abbreviate = TRUE)

# DATA CONVERSION: Ordering days of the week and months of the year 

# Make months and days of the week factors and order them as they are in a calendar
ntsb.data$Month = factor(ntsb.data$Month,levels=c("Jan", "Feb","Mar", "Apr","May",
                                                  "Jun","Jul","Aug","Sep","Oct", "Nov","Dec"), ordered=TRUE)


ntsb.data$Weekday = factor(ntsb.data$Weekday,levels=c("Sun","Mon","Tue",
                                                      "Wed","Thu","Fri","Sat"), ordered=TRUE)



# Step 5 Eliminate any rows (records) which has no date (Event.Date is NA)
elim.row = which(is.na(ntsb.data$Event.Date))
if (length(elim.row)>0) ntsb.data = ntsb.data[-elim.row,] # Eliminates all rows with only NA value for Event.Date

# Step 6: Changing blank, "N/A","Unknown", and similar responses to NA

# First, define  a vector of words or phrases that are to be replaced with NA
repl.words = c("", ".", "N/A", "UNK","Unknown","Unavailable")

# Note that the following checks all columns, event non-character one, but in the end,
# only character columns will have any non-zero values for replacement words
repl.list = apply(ntsb.data,2, function(x) which(x %in% repl.words))

# The number of times replacement words occur for each database variable (column), 
# is placed in a one-column data frame where the row names correspond to the 
# variable names the column has the number of times the replacement words occur
with.missing = as.data.frame(sapply(repl.list,length)) 
colnames(with.missing) = "Replacement.Words"

# Identify columns corresponding with character-based
# variables with at least one replacement word
with.missing.ndx = which(with.missing[,1]>0) 

# Replace replacement words with NAs in those columns containing
# one or more replacement words

for(i in 1:length(with.missing.ndx)){
        repl.vector=ntsb.data[,with.missing.ndx[i]]%in%repl.words 
        ntsb.data[repl.vector,with.missing.ndx[i]]=NA
} 


# Step 7: Specify city name, state abbreviation, and full state name
# for any location in the United states

# Will focus only on rows with a US city address entered, implying the following is true
# ntsb.data$Location is not NA and ntsb.data$Country=="United States"
city.locs = !(is.na(ntsb.data$Location)) & ntsb.data$Country=="United States"
# The following are the locations in the vector where this is true
city.locs.ndx = which(city.locs) 

# STATE CODES: Before adding full state names to data frame, must match
# two-letter state codes with full state names.

# First, create a vector of full state names by agmenting the built-in R 
# lists of names and abbreviations of the 50 states (state.name and state.abb) 
# with all other names and two-letter abbreviations used by the US Postal Service (USPS)

# Source is USPS Publication 28 - postal addressing standards 
# Located at http://pe.usps.gov/text/pub28/welcome.htm accessed 9 December 2015

extra.abb = c("AS", "DC", "FM","GU","MH", "MP", "PW", "PR", "VI", "GM", "AO","PO")
extra.name = c("American Samoa", "District of Columbia", "Federated States of Micronesia",
               "Guam", "Marshall Islands", "Northern Mariana Islands", "Palau",
               "Puerto Rico", "Virgin Islands", "Gulf of Mexico", "Atlantic Ocean",
               "Pacific Ocean")

# Now append them to the R-provided list of 50 state names and abbreviations
usps.abb = c(state.abb,extra.abb)
usps.state = c(state.name,extra.name)

# Next is to identify the number of commas in city identifiers
# This is because the NTSB output did not have separate variable for state
# States in NTSB output are two letter codes preceeded by a comma and a space

# These three vectors each initialized with a number of NA values equal to the
# number of rows to ensure that the final vector will be compatible with ntsb.data
comma.pos = rep(NA,nrow(ntsb.data)) # start with NA for number of commas for all Locations
city.vec = rep(NA,nrow(ntsb.data))
state.vec = rep(NA,nrow(ntsb.data))

for(x in 1:length(city.locs.ndx)){
        # Create a list that contains vector of comma positions
        comma.pos.list = gregexpr(",", ntsb.data$Location[city.locs.ndx[x]], fixed=TRUE)
        comma.pos.vec = comma.pos.list[[1]] # Vector of comma positinos
        comma.pos[x] = comma.pos.vec[length(comma.pos.vec)]
        
        # Get the length of the Location field character string
        num.chars = nchar(ntsb.data$Location[city.locs.ndx[x]]) 

        # Determine state code if location has enough characters for comma, space, and code
        if(comma.pos[x] >= 1 & num.chars >= 4){
                # Use last comma position to determine where split character string and find state code
                city.vec[city.locs.ndx[x]] =  substr(ntsb.data$Location[city.locs.ndx[x]], 1,(comma.pos[x]-1)) 
                state.vec[city.locs.ndx[x]] =  substr(ntsb.data$Location[city.locs.ndx[x]], (comma.pos[x]+2),num.chars)       
        } # End of if statement for creating city name and state abbreviation

} # End of process for finding US city names and state abbreviations     


# Initialize the full state name vector with a number of NA values equal to the
# number of rows to ensure that the final vector will be compatible with ntsb.data
state.full=rep(NA,nrow(ntsb.data))

for (i in 1:length(city.locs.ndx)){
        if(state.vec[i] %in% usps.abb) {
                state.full[i]=usps.state[grep(state.vec[i], usps.abb)]
        }
        # Erase city and two-letter state code if state code is not in usps.abb list
        if(!(state.vec[i] %in% usps.abb)) {
                city.vec[i]=NA
                state.vec[i]=NA
        }
}

# Can now add city and state abbrevications, and full state names to data frame
ntsb.data$City=city.vec
ntsb.data$State.code=state.vec
ntsb.data$State=state.full

# Step 8: Arrange the new columns in logical groupings

new.cols = c("Event.Id", "Investigation.Type", "Accident.Number",
             "Event.Date", "Year", "Month", "Day", "Weekday",
             "Location", "City", "State.code", "State", "Country", 
             "Airport.Code", "Airport.Name", "Latitude", "Longitude",
             "Injury.Severity", "Aircraft.Damage", "Aircraft.Category",
             "Registration.Number", "Make", "Model", "Amateur.Built",
             "Number.of.Engines", "Engine.Type", "FAR.Description", "Schedule",
             "Purpose.of.Flight", "Air.Carrier", "Total.Fatal.Injuries",
             "Total.Serious.Injuries", "Total.Minor.Injuries", "Total.Uninjured",
             "Weather.Condition", "Broad.Phase.of.Flight", "Report.Status",
             "Publication.Date")

ntsb.data = ntsb.data[,new.cols]
```
####Quick overview of data from the NTSB database####
Note that although the online NTSB database was intended for events from 1982 onwards, there were a handful of events from before 1982.

* Total number of records - `r paste(format(nrow(ntsb.data.raw), big.mark=","), sep="")`
* Number of records excluded - `r nrow(ntsb.data.raw)-nrow(ntsb.data)`
* Number of records processed - `r paste(format(nrow(ntsb.data), big.mark=","), sep="")`
* Number of processed records with a date before 1982 - `r paste(nrow(ntsb.data[ntsb.data$Year<1982,]))`
* Number of records with a US location - `r paste(format(nrow(ntsb.data[which(ntsb.data$Country=="United States"),]), big.mark=","), sep="")`
* Number of processed records inovlving fatalities `r paste(format(sum(ntsb.data$Total.Fatal.Injuries>=1, na.rm=TRUE), big.mark=","), sep="")`

####Example summary histograms####

```{r echo=FALSE}

# Vector of fatal events from 1982 onwards
fatal.vec = which(ntsb.data$Total.Fatal.Injuries>=1 & ntsb.data$Year>=1982)

# Data frame of all fatals from 1982 and beyond
fatal.df = ntsb.data[fatal.vec,]


# Histogram of top 15 states by number of events 1982-2015
barplot(sort(table(ntsb.data[ntsb.data$Year>=1982,]$State.code), 
        decreasing = TRUE)[1:15], col="dodgerblue",xlab="State", ylab="Events", 
        cex.names = 0.7, main="Top 15 states by number of events 1982-2015")

# Histogram of top 15 states by number of fatal events 1982-2015
barplot(sort(table(ntsb.data[fatal.vec,]$State.code), decreasing = TRUE)[1:15],
        col="dodgerblue",xlab="State", ylab="Fatal events",
        cex.names = 0.7, main="Top 15 states by number of fatal events: 1982-2015")

# Histogram events by day of the week
barplot(table(ntsb.data[ntsb.data$Year>=1982,]$Weekday), col="dodgerblue",
     xlab="Day", ylab="Events", main="Events by day of week: 1982 - 2015")

# Histogram of fatal events by day of the week
barplot(table(fatal.df$Weekday), col="dodgerblue", cex.names = 1.2,
        xlab="Day", ylab="Events", main="Fatal events by day of week: 1982 - 2015")


# Histogram of events by month of the year
barplot(table(ntsb.data[which(ntsb.data$Year>=1982),]$Month), col="dodgerblue", cex.names = 0.8,
        xlab="Month", ylab="Events", main="Events by month: 1982-2015")

# Histogram of fatal events by month of the year
barplot(table(fatal.df$Month), col="dodgerblue", cex.names = 0.8, xlab="Day",
        ylab="Fatal events", main="Fatal events by month: 1982-2015")

# Histogram of fatal events by year 1982 - 2015
barplot(table(ntsb.data[fatal.vec,]$Year), col="dodgerblue",
     xlab="Year", ylab="Fatal events", main="Fatal events by year: 1982 - 2015")

# -------
# Fatalites by year 1982 - 2015

# Do a tapply for sums by category then ensure it is table
death.table = as.table(tapply(fatal.df$Total.Fatal.Injuries, fatal.df$Year, sum))

barplot(death.table, col="dodgerblue",
     xlab="Year", ylab="Fatalities", main="Fatalities by year: 1982 - 2015")
# -------

```

###Discussion###
The NTSB database is one of the many data-related resources available on on some part of their site at www.ntsb.gov.  While many of them can be imported directly into spreadsheet or data analysis programs, users who have access to an R software installation may prefer to use R because of the program's many data analysis and data display options.

In addition, the R programs that are available in the links below may be adapted for other data resources that have a similar structure to the kinds of text and XML outputs provided by the NTSB.

###Resources###
NTSB Aviation Accident Database & Synopses
http://www.ntsb.gov/_layouts/ntsb.aviation/index.aspx

Data dictionary for the online NTSB database
http://www.ntsb.gov/_layouts/ntsb.aviation/AviationDownloadDataDictionary.aspx

Data dictionary for the output of the two AirSafe.com R programs
http://www.airsafe.com/analyze/ntsb-data-dictionary.pdf

R files used in this report 
https://github.com/airsafe/analyses/blob/master/ntsb.convert.txt.R (Text)
https://github.com/airsafe/analyses/blob/master/ntsb.convert.xml.R (XML)

Downloaded NTSB data used in this report
http://www.airsafe.com/analyze/AviationData.txt (Text)
http://www.airsafe.com/analyze/AviationData.xml (XML)

Processed NTSB data used in this report
http://www.airsafe.com/analyze/ntsb_data.csv

Data dictionary for the processed NTSB data
http://www.ntsb.gov/_layouts/ntsb.aviation/AviationDownloadDataDictionary.aspx

Converting NTSB database downloads (this report)
HTML - http://www.airsafe.com/analyze/ntsb.database.html
Rmd - https://github.com/airsafe/analyses/blob/master/ntsb.database.Rmd
RPubs - http://rpubs.com/airsafe/ntsb_database
